Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Sun leaves have anatomical, morphological, and physiological traits that reduce heat and water stress (Table `r table_leaf_traits`). They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a]. Higher stomatal and vein densities allow higher $g_s$ and facilitate higher transpiration [@zwienieckiHydraulicLimitationsImposed2004]. Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988]. These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but their greater thermal capacitance makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].

>>>>

Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to increased reflectance, lower wettability, and/or higher $g_b$, resulting in lower $T_{leaf}$-$T_{air}$ and/or reduced water stress (Table `r table_leaf_traits`). Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness, and higher trichome density [Fig. `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a]. Sun leaves tend to have higher photosynthetic rates per leaf area, related to higher nitrogen and chlorophyll concentrations per leaf area and lower intercellular CO~2~ concentrations. Further, sun leaves tend to have higher stomatal and vein densities would enable their higher transpiration rates, and that would facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004]. Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and, in addition to higher droplet contact angles, would have reduced wettability. Leaves in high light environments also exhibit greater leaf packing and clumping. In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance, larger leaf size and lower transpiration--makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies, where light is comparatively homogeneous, leaf traits may be more strongly related to maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
